\documentclass[10pt,titlepage,twoside,a4paper]{report}
\usepackage[left=3cm,right=3cm,top=3.5cm,bottom=2.5cm]{geometry}
%\usepackage[icelandic]{babel}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{graphicx}
\usepackage{parskip}
\usepackage{url}
\usepackage{apacite}
\usepackage{amssymb}
\usepackage[tight,footnotesize]{subfigure}

\usepackage{color}

\setcounter{secnumdepth}{3}
\setcounter{tocdepth}{1}

\author{Kristj\'{a}n R\'{u}narsson \and Benedikt Kristinsson\\\and Supervisor: Kristj\'{a}n Valur J\'{o}nsson}
\title{NSN Project Final Report\\\textit{TSense}: A Trusted Sensor and Support Infrastructure}
\date{20.09.2010}

\begin{document}

\maketitle
%\begin{titlepage}
%\end{titlepage}
\newpage
\ \ \ \\
\newpage

\pagenumbering{roman}
\setcounter{page}{1}

\tableofcontents
\pagestyle{empty}
\newpage
\ \ \ \\
\newpage
\pagebreak

\begin{abstract}
Distributed measurement and monitoring systems are vital to the functioning of many critical systems and are due to be ever more important as systems grow in size and complexity. This raises questions regarding the trustworthiness of the measurement processes, critical for making prudent manual or automatic control decisions. 

In this work, we address one particular aspect of secure distributed measurement systems. We describe the design and construction of a trusted sensor prototype, a small tamper-proof device which gives cryptographic guarantees of data authenticity, and optionally, confidentiality. Further, we describe the design and construction of a comprehensive client/server based measurement system, using the trusted sensor to provide end-to-end trust guarantees. 


%
Our prototype system was successfully implemented and tested on a small-scale test system of four separate network connected nodes.
\end{abstract}

\renewcommand{\abstractname}{{\'{U}}tdr\'{a}ttur}
\begin{abstract}
Dreifð nettengd mælikerfi eru stór þáttur í rekstri margra mikilvægra kerfa og munu væntanlega verða enn nauðsynlegri eftir því sem kerfi stækka og verða flóknari. Þessi þróun vekur spurningar varðand öryggi slíkra mæliferla og sérstaklega það traust sem hægt er að bera til þeirra. Traustar niðurstöður eru mikilvægur þáttur í því að hægt sé að taka réttar handvirkar eða sjálfvirkar stjórnákvarðanir.

Við skoðum einn þátt öruggra dreifðra mælineta í þessu verkefni. Við lýsum hönnun og smíði trausts skynara, sem er lítið tæki í óopnanlegri umlykju (e. tamper-proof) sem gefur kriptografíska tryggingu fyrir gagnatrausti og duld. Einnig lýsum við hönnun og smíði heildstæðs biðlara-miðlara mælinets sem  notar trausta skynjarann til að gefa viðtakanda áreiðanlegar niðurstöður.
%
Frumgerð okkar og stoðkerfi var smíðuð og sett upp á litlu prófunarkerfi með fjórum nettengdum tækjum.
\end{abstract}

\pagenumbering{arabic}
\setcounter{page}{1}
\pagestyle{headings}

\input{intro}
\input{method}
\input{results}
\input{discussion}

\bibliographystyle{apacite}
\bibliography{tsense}

\end{document}
